---
title: "Final Project 2_Ocean Health Index Analysis"
author: "Kyria Santa"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

## Data Set Description

This analysis uses the Ocean Health Index (OHI) dataset, which provides annua assessments of ocean health for over 200 coastal countries and territories from 2012 to 2024.The dataset includes: 

* REF_AREA: Country/region code ("USA")

* REF_AREA_LABEL: Full country/region name ("United States")

* Yearly OHI scores: Numerical values (0-100 scale) for years 2012 through 2024

The OHI score represents the overall health of marine systems, integrating ecological, social, and economic lenses. Higher OHIs indicate better ocean health.

## Analysis Objectives

1. Data Prep: Load and reshape the OHI data from wide to tidy format

2. Descriptive Analysis: Identify top-performing and bottom-performing countries

3. Trend Analysis: Visualize global and regional trends over time

4. Comparative Analysis: Examine specific countries of interest

# Data Loading
```{r}
# Installing and Loading Necessary Libraries
install.packages("rmarkdown")
install.packages("tidyverse")
install.packages("knitr")
install.packages("ggplot2")
install.packages("kableExtra")
library("tidyverse")
library("knitr")
library("ggplot2")
library("kableExtra")
library("rmarkdown")


# Reading my CSV File
ohi_data <- read.csv(file.path("..", "data", "ocean_health_index.csv"))

# Viewing my Data
head(ohi_data)
```

# Data Prep
```{r}
# Reshaping from wide to "tidy" format
ohi_tidy <- ohi_data %>%
  pivot_longer(
    cols = X2012:X2024,
    names_to = "Year",
    values_to = "OHI_Score"
  ) %>%
  mutate(
    Year = as.numeric(gsub("X", "", Year)),
    OHI_Score = as.numeric(OHI_Score)
  ) %>%
  arrange(REF_AREA, Year)

# Verifying this worked
head(ohi_tidy)

# Checking for Missing Values
cat("\nMissing values in Year column:", sum(is.na(ohi_tidy$Year)))
cat("\nMissing values in OHI_Score column:", sum(is.na(ohi_tidy$OHI_Score)))
```


# Analysis Results

## Descriptive Statistics Table

```{r}
# Calculating Summary Statistics by Year
yearly_summary <- ohi_tidy %>%
  group_by(Year) %>%
  summarize(
    Mean_OHI = mean(OHI_Score, na.rm = TRUE),
    Median_OHI = median(OHI_Score, na.rm = TRUE),
    SD_OHI = sd(OHI_Score, na.rm = TRUE),
    Min_OHI = min(OHI_Score, na.rm = TRUE),
    Max_OHI = max(OHI_Score, na.rm = TRUE),
    N_Countries = n_distinct(REF_AREA)
  ) %>%
  round(2)

# Creating a table
yearly_summary %>% 
  kable(
    caption = "Annual Summary Statistics of Ocean Health Index (2012-2024)",
    col.names = c("Year", "Mean", "Median", "Std Dev", "Min", "Max", "N Countries")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```
# Table Description 

The table above presents summary statistics of the Ocean Health Index across all countries for each year from 2012 to 2024. Key observations include: 

* Global averages range from about 69 to 71 points, indicating moderate ocean health overall

* Variability is consistent across years (st devs around 6-7 points)

* Complete data: All years include scores for the same number of countries (194), suggesting no missing yearly data

* 2023-2024 dip: The mean scores show a noticeable decrease in the most recent years, suggesting potential global declines in ocean health 

# Top and Bottom Performing Countries (2024)

```{r}
# Finding top and bottom performers in 2024
latest_year <- ohi_tidy %>%
  filter(Year == 2024) %>%
  arrange(desc(OHI_Score))

top_10 <- latest_year %>%
  slice_head(n = 10) %>%
  select(Country = REF_AREA_LABEL, `2024 Score` = OHI_Score)

bottom_10 <- latest_year %>%
  slice_tail(n = 10) %>%
  select(Country = REF_AREA_LABEL, `2024 Score` = OHI_Score)

# Creating a combined comparison table
comparison_df <- data.frame(
  Rank = c(paste0("Top ", 1:10), paste0("Bottom ", 1:10)),
  Country = c(top_10$Country, bottom_10$Country),
  Score = c(top_10$`2024 Score`, bottom_10$`2024 Score`)
)

comparison_df %>%
  kable(
    caption = "Top and Bottom Performing Countries in Ocean Health (2024)",
    col.names = c("Rank", "Country", "OHI Score")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

# Data Visualization

## Figure 1: Global OHI Trends Over Time
```{r}
# Creating a plot showing global trends
global_avg <- ohi_tidy %>%
  group_by(Year) %>%
  summarize(
    Global_Mean = mean(OHI_Score, na.rm = TRUE),
    Lower_CI = Global_Mean - sd(OHI_Score, na.rm = TRUE)/sqrt(n()),
    Upper_CI = Global_Mean + sd(OHI_Score, na.rm = TRUE)/sqrt(n())
  )

ggplot(global_avg, aes(x = Year, y = Global_Mean)) +
  geom_line(color = "steelblue", linewidth = 1.5) +
  geom_ribbon(aes(ymin = Lower_CI, ymax = Upper_CI),
              alpha = 0.2, fill = "steelblue") +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Figure 1: Global Ocean Health Index Trends (2012-2024)",
    subtitle = "Mean Scores with 95% Confidence Intervals",
    x = "Year",
    y = "Ocean Health Index Score",
    caption = "Data Source: Ocean Health Index (Kaggle)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.text = element_text(color = "black", size = 11),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(size = 10)
  ) + 
  scale_x_continuous(breaks = seq(2012, 2024, 2)) +
  scale_y_continuous(limits = c(65, 75), breaks = seq(65, 75, 1))
```

# Figure Description

Figure 1 shows the global average Ocean Health Index (OHI) from 2012 to 2024 with 95% confidence intervals. Key insights:

1. Overall Trend: Global ocean health showed improvement from 2012 to 2020, reaching a peak around 71.5

2. Recent Decline: A noticeable dip occurs between 2023-2024, suggesting potential environmental pressures or human influence

3. Statistical Significance: The confidence intervals narrow over time, indicating more precise estimates in recent years

4. Variability: The shaded area represents uncertainty; wider intervals in earlier years suggest more variable measurements 


# Figure 2: Country-Specific Trends Comparison
```{r}
# Selecting countries for comparison
selected_countries <- c("United States", "India", "China", "Sudan", "Syrian Arab Republic", "World")

# Calculating global average for comparison
global_trend <- ohi_tidy %>%
  group_by(Year) %>%
  summarize(OHI_Score = mean(OHI_Score, na.rm = TRUE)) %>%
  mutate(REF_AREA_LABEL = "World")

# Getting country trends
country_trends <- ohi_tidy %>%
  filter(REF_AREA_LABEL %in% selected_countries) %>%
  select(REF_AREA_LABEL, Year, OHI_Score)

# Combining with global average
comparison_data <- bind_rows(country_trends, global_trend) %>%
  mutate(Country = factor(REF_AREA_LABEL, levels = selected_countries))

ggplot(comparison_data, aes(x = Year, y = OHI_Score, color = Country, group = Country)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Figure 2: Ocean Health Index Trends: Selected Countries vs. Global Average",
    x = "Year",
    y = "Ocean Health Index Score",
    color = "Country/Region",
    caption = "Note: 'World' represents the global average score"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 11),
    axis.text = element_text(color = "black", size = 11),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(size = 10)
  ) +
  scale_x_continuous(breaks = seq(2012, 2024, 2)) +
  scale_color_brewer(palette = "Set1") +
  guides(color = guide_legend(nrow = 1))
```

# Figure Description

Figure 2 compares OHI trends for five selected countries to the global average. Observations:

1. Performance Variation: Countries shows diverse trajectories, with some consistently above (United States) or below (Syrian Arab Republic) the global average

2. Convergence Patterns: Most countries show the 2023-2024 dip, suggesting a shared global occurrence

3. Country-Specific Observations:
* The Syrian Arab Republic shows a sharp improvement form 2019-2020, holds this improvement until 2021, but faces a sharp decline from 2022-2023

* China shows steady improvement over time

* United States shows a steady decline over time 

4. Benchmark Use: The "World" line provides a useful reference for evaluating individual country performance

# Conclusion

This analysis demonstrates the use of the Ocean Health Index dataset for tracking marine ecosystem health over time. The reproducible pipeline developed here successfully:

1. Processed raw data from wide to tidy format, adjusting for the X-prefixed year columns

2. Generated descriptive statistics and formatted tables

3. Created informative visualizations with proper formatting 

4. Identified key trends, including the global peak in 2022 and following decline

This R Markdown file ensures reproducibility, with all code, data transformations, and analyses documented in a single file. In the future, I could extend this analysis by adding more environmental or socioeconomic variables to better understand the drivers of ocean health changes. 